Spatiotemporal Multi-Resolution Approximation of the Amari Type Neural Field Model

### 2012-01-31 Multiresolution_framework/ltx/manuscript.tex
minor edits

By assuming a linear activation function of the form
\begin{equation}
	f(v_t(\mathbf{r})) = \varsigma v_t(\mathbf{r})
\end{equation}
the state transition equation~\eqref{eq:ApproxDiscreteTimeModel4} simplifies to
\begin{equation}\label{eq:ApproxDiscreteTimeModel_Linear}
	\mathbf{x}_{t+1} = 
	\xi \mathbf{x}_t + 
	\varsigma T_s \mathbf{\Lambda}_{x}^{-1} \int_{\Omega}\boldsymbol\mu\left(\mathbf{r}\right)\int_\Omega { 
	    \boldsymbol\theta^\top\boldsymbol\lambda\left(\mathbf{r}-\mathbf{r}'\right)
	    \boldsymbol\mu^\top\left(\mathbf{r}'\right) 
	\, \mathrm{d}\mathbf{r}'\mathrm{d}\mathbf{r}} \mathbf{x}_t
	+ \mathbf{w}_t.
\end{equation}
Now we can further simplify the system by defining
\begin{align}
	\label{eq:Lambdatheta}
	 \mathbf{\Lambda}_{\theta} &\triangleq \int_{\Omega}\boldsymbol\mu\left(r\right) \int_\Omega { 
		   \boldsymbol\theta^\top\boldsymbol\lambda\left(r-r'\right)
		    \boldsymbol\mu^\top\left(r\right)\ \mathrm{d}r'\mathrm{d}r}
\end{align}
to give the state transition equation
\begin{align}\label{eq:StateEquation}
 \mathbf x_{t+1} &=\mathbf A(\boldsymbol \theta) \mathbf x_t+ \mathbf w_t\\
\label{eq:A_theta}
 \mathbf A(\boldsymbol \theta) &= T_s\varsigma\mathbf{\Lambda}_{x}^{-1}\mathbf{\Lambda}_{\theta}+\xi\mathbf I.
\end{align} 
The problem of joint state and parameter estimation for the state-space model given by \eqref{eq:StateEquation} and \eqref{eq:ReducedObservationEquation} can be formulated as 
\begin{equation}
	{\hat{\mathbf X},\hat{\boldsymbol\theta}}=\arg\max_{\mathbf X,\boldsymbol\theta}p(\mathbf Y;\boldsymbol\theta),
 \end{equation}  
where $\mathbf X$ and $\mathbf Y$ denote the entire sequences of the states and observations respectively, $\boldsymbol\theta$ is the parameter set and $p(\mathbf Y;\boldsymbol\theta)$ denotes the probability density function of the observations, $\mathbf Y$, parameterised  by $\boldsymbol\theta$, known as the likelihood function. This problem cannot be solved separately as for the state estimation the knowledge of the system's parameters is required and for the parameter estimation the states of the system need to be known. One approach is to iteratively estimate the states and parameters of the system and monitor the convergence of the algorithm. A natural framework to perform this is the well known Expectation Maximisation (EM) algorithm \cite{Dempster1977,Shumway2000} to infer both the parameter set and the states from the observations corresponding to the connectivity kernel, the neural field and electrophysiological data in our application. There are two main applications of the EM algorithm:  first is when the observations are incomplete due to a faulty observation process or existing limitations and the second  when the direct maximisation of the likelihood function is difficult but can be simplified by assuming the existence of an additional however missing (or hidden) variables \cite{Bilmes1998}, the weights of the field basis functions in this case. The EM algorithm, when used in the latter context \cite{Dewar2009}, yields the maximum likelihood kernel estimate, i.e.
\begin{equation}
	\boldsymbol\theta_{\text{ML}}=\arg\max_{\boldsymbol\theta}~p(\mathbf Y;\boldsymbol\theta),
 \end{equation}   
and the posterior distribution of the field over time.

@@

In summary, to solve the joint state and parameter estimation for linear state-space models one can adapt the EM algorithm, solving the smoothing problem for the E-step and constructing and maximising the lower bound in the M-step. A schematic representation of the EM algorithm is represented in Figure~\ref{fig:EMBlockDiagram}.  
\begin{figure}[!t]
\centering
\includegraphics[scale=1]{./Graph/EMBlockDiagram.pdf}
\caption{{\bf Schematic of the EM algorithm}. The RTS smoother (E-step) and the maximisation of the lower bound, $\mathcal Q(\boldsymbol \theta,\boldsymbol\theta')$, (M-step) are shown. The dashed box shows the iterative two step algorithm once it is initialised.}
\label{fig:EMBlockDiagram}
\end{figure}
  \subsection{E-step}
  As mentioned earlier the E-step for state-space models is equivalent to the smoothing problem. We use the RTS smoother in order to infer the states of the system. The RTS smoother uses the Kalman filter \cite{Kalman1960} in forward pass along withe a backward iteration to calculate its outputs: smoothed states $\hat{\mathbf x}^b_t$ and covariances, $\mathbf P^b_t=\mathrm{cov}(\mathbf{x}_t)$. Following \cite{Gibsona2005}, an additional backwards recursion is also included (step 6 of the RTS algorithm)	to calculate cross-covariance matrix, $\mathbf M_t=\mathrm{cov}(\mathbf{x}_{t},\mathbf{x}_{t+1})$, which is also required in forming the lower bound to be maximised in the M-step. Note that the following quantities from the forward iteration (filtering) should be stored when implementing the RTS smoother: predicted and corrected estimates for both state estimates ($\mathbf{x}_t^{f-}$, $\mathbf{x}_t^{f}$) and covariance matrices ($\mathbf P_t^{f-}$,$\mathbf P_t^f$). Also the final value of the Kalman gain, $\mathcal K_T$, should be stored in order to initialise the backward cross-covariance computation. 
 \subsection{M-step}
It is rather straightforward to show that 
\begin{equation}
	\mathcal Q(\boldsymbol \theta,\boldsymbol\theta')= \mathbf E_{\boldsymbol \theta'}\left[2\ln p(\mathbf X,\mathbf Y;\boldsymbol \theta)\right], 
\end{equation}
is a lower bound on the log-likelihood function \cite{Bishop2006} (see Appendix...)  where $p(\mathbf X,\mathbf Y;\boldsymbol \theta)$ is the joint probability distribution of the states and the observations and where $ \mathbf E_{\boldsymbol \theta'}\left[\cdot\right] $ denotes expectation taken with respect to the marginal distribution of the states conditioned on the observed field and the current estimate of the parameter set, $\boldsymbol\theta'$ i.e.  $p(\mathbf X\mid\mathbf Y;\boldsymbol \theta'$). Note that this is actually the smoother output.

@@

The matrices $\boldsymbol\Xi_0$ and $\boldsymbol\Xi_1$ are calculated using the RTS smoother outputs: $\hat{\mathbf x}_t$, covariance, $\mathbf P_t^b$, and cross-covariance matrix, $\mathbf M_t$ \cite{Gibsona2005}, (see Appendix~ for more details)

@@

\subsection{Lower bound on likelihood function}\label{ap:MStep}
A derivation of the lower bound on the likelihood function that is based on \citet{Bishop2006} is provided in this Appendix. We want to maximise the likelihood function which is difficult to perform directly
\begin{equation}
 p(\mathbf Y;\boldsymbol\theta)=\int_{\mathbf X}p(\mathbf X,\mathbf Y;\boldsymbol\theta)d\mathbf X.
\end{equation}   
Maximising the likelihood in terms of the parameter set is equivalent to the log-likelihood maximisation, as both the likelihood and log-likelihood functions share the same maximiser due to the monotonicity of the logarithm function. For any choice of distribution defined over the latent variables, $q(\mathbf X)$, the following decomposition holds
\begin{equation}\label{eq:EM-loglikelihooddecomposition}
 \ln~p(\mathbf Y;\boldsymbol\theta)=\mathcal{L}(q,\boldsymbol\theta)+\mathrm{KL}(q\parallel p),
\end{equation}
where
\begin{align}
 \mathcal{L}(q,\boldsymbol\theta)&=\int_{\mathbf X} q(\mathbf X)\ln\left\lbrace \frac{p(\mathbf X,\mathbf Y;\boldsymbol\theta)}{q(\mathbf X)} \right\rbrace d\mathbf X \label{eq:EM-decomposition1}\\
\mathrm{KL}(q\parallel p)&=-\int_{\mathbf X}q(\mathbf X)\ln\left\lbrace \frac{p(\mathbf X \mid \mathbf Y;\boldsymbol\theta)}{q(\mathbf X)} \right\rbrace d\mathbf X. \label{eq:EM-decomposition2}
\end{align}
Equation \eqref{eq:EM-loglikelihooddecomposition} can be verified by applying the probability product rule to the joint distribution $p(\mathbf X,\mathbf Y;\boldsymbol\theta)$ in \eqref{eq:EM-decomposition1} giving
\begin{align}
	\mathcal{L}(q,\boldsymbol\theta)+\mathrm{KL}(q\parallel p)&= \int_{\mathbf X} q(\mathbf X)\ln\left\lbrace \frac{p(\mathbf X \mid \mathbf Y;\boldsymbol\theta)p(\mathbf Y;\boldsymbol\theta)}{q(\mathbf X)} \right\rbrace d\mathbf X \nonumber\\
	&- \int_{\mathbf X}q(\mathbf X)\ln\left\lbrace \frac{p(\mathbf X \mid \mathbf Y;\boldsymbol\theta)}{q(\mathbf X)} \right\rbrace d\mathbf X\nonumber\\
	&= \int_{\mathbf X}q(\mathbf X)\ln p(\mathbf Y; \boldsymbol\theta)d\mathbf X.
\end{align}
Noting  $q(\mathbf X)$ is a normalised distribution that integrates to one gives the log-likelihood function. Note \eqref{eq:EM-decomposition2} is the Kullback-Liebler distance between distributions $p(\mathbf X \mid \mathbf Y;\boldsymbol\theta)$ and $q(\mathbf X)$, therefore,
\begin{equation}
	\mathrm{KL}(q\parallel p)\ge 0.
\end{equation}   
The equality holds if and only if $q(\mathbf X)=p(\mathbf X \mid \mathbf Y;\boldsymbol\theta)$. It follows from equation \eqref{eq:EM-loglikelihooddecomposition}, that the quantity $\mathcal{L}(q,\boldsymbol\theta)$ is a lower bound on the log-likelihood function, i.e.
\begin{equation}
 \mathcal{L}(q,\boldsymbol\theta)\le \ln~p(\mathbf Y\mid\boldsymbol\theta)
	\end{equation}

The EM algorithm alternates between maximising the lower bound with respect to the distribution $q(\mathbf X)$ and the parameter set $\boldsymbol\theta$, respectively, while holding the other fixed. Suppose that the current estimate of the parameter set is $\boldsymbol\theta'$, in the E-step the lower bound is maximised with respect to the distribution $q(\mathbf X)$ while $\boldsymbol\theta'$ is fixed. Since the value of $p(\mathbf Y;\boldsymbol\theta')$ is independent of $q(\mathbf X)$ the maximum of the lower bound occurs when the KL distance disappears or in other words when $q(\mathbf X)$ is exactly the conditional distribution $p(\mathbf X \mid \mathbf Y;\boldsymbol\theta)$. At this point, the lower bound becomes equal to the log-likelihood function. Note in the state-space model the E-step is equivalent to the smoothing problem. The maximum in the M-step is obtained by maximising the lower bound with respect to the parameter set while the distribution $q(\mathbf X)$ is held fixed. Since the distribution $q(\mathbf X)$ is computed using the old parameter estimates and kept fixed during the M-step, it is not equal to $p(\mathbf X \mid \mathbf Y;\boldsymbol\theta)$, resulting in a non-zero KL distance. Therefore, the growth of the log-likelihood function is greater than the growth of the lower bound. Substituting $p(\mathbf X \mid \mathbf Y;\boldsymbol\theta')$ for $q(\mathbf X)$ in \eqref{eq:EM-decomposition1} gives
\begin{align}
 \mathcal{L}(q,\boldsymbol\theta)&=\int_{\mathbf X} p(\mathbf X\mid\mathbf Y;\boldsymbol\theta')\ln p(\mathbf X,\mathbf Y;\boldsymbol\theta)~d\mathbf X-\int_{\mathbf X} p(\mathbf X\mid\mathbf Y;\boldsymbol\theta')\ln p(\mathbf X\mid\mathbf Y;\boldsymbol\theta')~d\mathbf X \nonumber \\
&= \mathbf E_{\boldsymbol\theta'}\left[ \ln p(\mathbf X,\mathbf Y;\boldsymbol\theta)\right] +\mathrm{const.}\nonumber\\
&=\mathcal{Q}(\boldsymbol\theta,\boldsymbol\theta')+\mathrm{const.}~,\label{eq:EM-ComDaLogLike}
\end{align}
where $ \mathbf E_{\boldsymbol \theta'}\left[ \cdot\right] $ denotes expectation taken with respect to the marginal distribution of the states conditioned on the observed field and the current estimate of the parameter set, i.e.  $p(\mathbf X\mid\mathbf Y;\boldsymbol \theta'$). The constant term is the negative entropy of $q(\mathbf X)$ and is independent of $\boldsymbol\theta$.   It becomes clear from \eqref{eq:EM-ComDaLogLike} that, the M-step essentially maximises the expectation of the complete data log-likelihood. 
\section{Calculation of $\boldsymbol\Xi_{0}$ and $\boldsymbol\Xi_{1}$}
The expressions for $\boldsymbol\Xi_{0}$ and $\boldsymbol\Xi_{1}$ in terms of smoother outputs can be found in \cite{Shumway2000}, here we present the derivation for completeness.  The expression for  $\boldsymbol\Xi_{0}$ as defined in equation~\eqref{eq:Xi0} of the main text is given by
   \begin{equation}\label{eq:appXi0}
	 \boldsymbol\Xi_0=\mathbf E_{\theta'}\left[\sum_{t=0}^{T-1}\mathbf x_t\mathbf x_{t+1}^\top\right].
	\end{equation}                                                                                                
	We start by expanding the equation \eqref{eq:appXi0}, giving

>>>>

By assuming a linear activation function of the form
\begin{equation}
	f(v_t(\mathbf{r})) = \varsigma v_t(\mathbf{r})
\end{equation}
the state transition equation~\eqref{eq:ApproxDiscreteTimeModel4} simplifies to
\begin{equation}\label{eq:ApproxDiscreteTimeModel_Linear}
	\mathbf{x}_{t+1} = 
	\xi \mathbf{x}_t + 
	\varsigma T_s \mathbf{\Lambda}_{x}^{-1} \int_{\Omega}\boldsymbol\mu\left(\mathbf{r}\right)\int_\Omega { 
	    \boldsymbol\theta^\top\boldsymbol\lambda\left(\mathbf{r}-\mathbf{r}'\right)
	    \boldsymbol\mu^\top\left(\mathbf{r}'\right) 
	\, \mathrm{d}\mathbf{r}'\mathrm{d}\mathbf{r}} \mathbf{x}_t
	+ \mathbf{w}_t.
\end{equation}
Now we can further simplify the system by defining
\begin{align}
	\label{eq:Lambdatheta}
	 \mathbf{\Lambda}_{\theta} &\triangleq \int_{\Omega}\boldsymbol\mu\left(r\right) \int_\Omega { 
		   \boldsymbol\theta^\top\boldsymbol\lambda\left(r-r'\right)
		    \boldsymbol\mu^\top\left(r\right)\ \mathrm{d}r'\mathrm{d}r}
\end{align}
to give the state transition equation
\begin{align}\label{eq:StateEquation}
 \mathbf x_{t+1} &=\mathbf A(\boldsymbol \theta) \mathbf x_t+ \mathbf w_t\\
\label{eq:A_theta}
 \mathbf A(\boldsymbol \theta) &= T_s\varsigma\mathbf{\Lambda}_{x}^{-1}\mathbf{\Lambda}_{\theta}+\xi\mathbf I.
\end{align} 
The problem of joint state and parameter estimation for the state-space model given by \eqref{eq:StateEquation} and \eqref{eq:ReducedObservationEquation} can be formulated as 
\begin{equation}
	{\hat{\mathbf X},\hat{\boldsymbol\theta}}=\arg\max_{\mathbf X,\boldsymbol\theta}p(\mathbf Y;\boldsymbol\theta),
 \end{equation}  
where $\mathbf X$ and $\mathbf Y$ denote the entire sequences of the states and observations respectively, $\boldsymbol\theta$ is the parameter set and $p(\mathbf Y;\boldsymbol\theta)$ denotes the probability density function of the observations parameterised  by $\boldsymbol\theta$, known as the likelihood function. This problem cannot be solved separately as for the state estimation the knowledge of the system's parameters is required and for the parameter estimation the states of the system need to be known. One approach is to iteratively estimate the states and parameters of the system and monitor the convergence of the algorithm. A natural framework to perform this is the well known Expectation Maximisation (EM) algorithm \cite{Dempster1977,Shumway2000} to infer both the parameter set and the states from the observations corresponding to the connectivity kernel, the neural field and electrophysiological data in our application. There are two main applications of the EM algorithm:  first is when the observations are incomplete due to a faulty observation process or existing limitations and the second  when the direct maximisation of the likelihood function is difficult but can be simplified by assuming the existence of an additional however missing (or hidden) variables \cite{Bilmes1998}, the weights of the field basis functions in this case. The EM algorithm, when used in the latter context \cite{Dewar2009}, yields the maximum likelihood kernel estimate, i.e.
\begin{equation}
	\boldsymbol\theta_{\text{ML}}=\arg\max_{\boldsymbol\theta}~p(\mathbf Y;\boldsymbol\theta),
 \end{equation}   
and the posterior distribution of the field over time.

@@

In summary, to solve the joint state and parameter estimation for linear state-space models one can adapt the EM algorithm, solving the smoothing problem for the E-step and constructing and maximising the lower bound in the M-step. A schematic representation of the EM algorithm is represented in Figure~\ref{fig:EMBlockDiagram}.  
\begin{figure}[!t]
\centering
\includegraphics[scale=1]{./Graph/EMBlockDiagram.pdf}
\caption{{\bf Schematic of the EM algorithm}. The RTS smoother (E-step) and the maximisation of the lower bound, $\mathcal Q(\boldsymbol \theta,\boldsymbol\theta')$, (M-step) are shown. The dashed box shows the iterative two step algorithm once it is initialised.}
\label{fig:EMBlockDiagram}
\end{figure}
  \subsection{E-step}
  As mentioned earlier the E-step for state-space models is equivalent to the smoothing problem. We use the RTS smoother in order to infer the states of the system. The RTS smoother uses the Kalman filter \cite{Kalman1960} in forward pass along withe a backward iteration to calculate its outputs: smoothed states $\hat{\mathbf x}^b_t$ and covariances, $\mathbf P^b_t=\mathrm{cov}(\mathbf{x}_t)$. Following \cite{Gibsona2005}, an additional backwards recursion is also included (step 6 of the RTS algorithm)	to calculate cross-covariance matrix, $\mathbf M_t=\mathrm{cov}(\mathbf{x}_{t},\mathbf{x}_{t+1})$, which is also required in forming the lower bound to be maximised in the M-step. Note that the following quantities from the forward iteration (filtering) should be stored when implementing the RTS smoother: predicted and corrected estimates for both state estimates ($\mathbf{x}_t^{f-}$, $\mathbf{x}_t^{f}$) and covariance matrices ($\mathbf P_t^{f-}$,$\mathbf P_t^f$). Also the final value of the Kalman gain, $\mathcal K_T$, should be stored in order to initialise the backward cross-covariance computation. 
 \subsection{M-step}
It is rather straightforward to show that 
\begin{equation}
	\mathcal Q(\boldsymbol \theta,\boldsymbol\theta')= \mathbf E_{\boldsymbol \theta'}\left[2\ln p(\mathbf X,\mathbf Y;\boldsymbol \theta)\right], 
\end{equation}
is a lower bound on the log-likelihood function \cite{Bishop2006} (see Appendix~\ref{ap:Lowerbound} )  where $p(\mathbf X,\mathbf Y;\boldsymbol \theta)$ is the joint probability distribution of the states and the observations and where $ \mathbf E_{\boldsymbol \theta'}\left[\cdot\right] $ denotes expectation taken with respect to the marginal distribution of the states conditioned on the observed field and the current estimate of the parameter set, $\boldsymbol\theta'$ i.e.  $p(\mathbf X\mid\mathbf Y;\boldsymbol \theta'$). Note that this is actually the smoother output.

@@

The matrices $\boldsymbol\Xi_0$ and $\boldsymbol\Xi_1$ are calculated using the RTS smoother outputs: $\hat{\mathbf x}_t$, covariance, $\mathbf P_t^b$, and cross-covariance matrix, $\mathbf M_t$ \cite{Gibsona2005}, (see Appendix~\ref{ap:Xiderivation} for more details)

@@

\subsection{Lower bound on likelihood function}\label{ap:Lowerbound}
A derivation of the lower bound on the likelihood function that is based on \citet{Bishop2006} is provided in this Appendix. We want to maximise the likelihood function which is difficult to perform directly
\begin{equation}
 p(\mathbf Y;\boldsymbol\theta)=\int_{\mathbf X}p(\mathbf X,\mathbf Y;\boldsymbol\theta)d\mathbf X.
\end{equation}   
Maximising the likelihood in terms of the parameter set is equivalent to the log-likelihood maximisation, as both the likelihood and log-likelihood functions share the same maximiser due to the monotonicity of the logarithm function. For any choice of distribution defined over the latent variables, $q(\mathbf X)$, the following decomposition holds
\begin{equation}\label{eq:EM-loglikelihooddecomposition}
 \ln~p(\mathbf Y;\boldsymbol\theta)=\mathcal{L}(q,\boldsymbol\theta)+\mathrm{KL}(q\parallel p),
\end{equation}
where
\begin{align}
 \mathcal{L}(q,\boldsymbol\theta)&=\int_{\mathbf X} q(\mathbf X)\ln\left\lbrace \frac{p(\mathbf X,\mathbf Y;\boldsymbol\theta)}{q(\mathbf X)} \right\rbrace d\mathbf X \label{eq:EM-decomposition1}\\
\mathrm{KL}(q\parallel p)&=-\int_{\mathbf X}q(\mathbf X)\ln\left\lbrace \frac{p(\mathbf X \mid \mathbf Y;\boldsymbol\theta)}{q(\mathbf X)} \right\rbrace d\mathbf X. \label{eq:EM-decomposition2}
\end{align}
Equation \eqref{eq:EM-loglikelihooddecomposition} can be verified by applying the probability product rule to the joint distribution $p(\mathbf X,\mathbf Y;\boldsymbol\theta)$ in \eqref{eq:EM-decomposition1} giving
\begin{align}
	\mathcal{L}(q,\boldsymbol\theta)+\mathrm{KL}(q\parallel p)&= \int_{\mathbf X} q(\mathbf X)\ln\left\lbrace \frac{p(\mathbf X \mid \mathbf Y;\boldsymbol\theta)p(\mathbf Y;\boldsymbol\theta)}{q(\mathbf X)} \right\rbrace d\mathbf X \nonumber\\
	&- \int_{\mathbf X}q(\mathbf X)\ln\left\lbrace \frac{p(\mathbf X \mid \mathbf Y;\boldsymbol\theta)}{q(\mathbf X)} \right\rbrace d\mathbf X\nonumber\\
	&= \int_{\mathbf X}q(\mathbf X)\ln p(\mathbf Y; \boldsymbol\theta)d\mathbf X.
\end{align}
Noting  $q(\mathbf X)$ is a normalised distribution that integrates to one gives the log-likelihood function. Note \eqref{eq:EM-decomposition2} is the Kullback-Liebler distance between distributions $p(\mathbf X \mid \mathbf Y;\boldsymbol\theta)$ and $q(\mathbf X)$, therefore,
\begin{equation}
	\mathrm{KL}(q\parallel p)\ge 0.
\end{equation}   
The equality holds if and only if $q(\mathbf X)=p(\mathbf X \mid \mathbf Y;\boldsymbol\theta)$. It follows from equation \eqref{eq:EM-loglikelihooddecomposition}, that the quantity $\mathcal{L}(q,\boldsymbol\theta)$ is a lower bound on the log-likelihood function, i.e.
\begin{equation}
 \mathcal{L}(q,\boldsymbol\theta)\le \ln~p(\mathbf Y\mid\boldsymbol\theta)
	\end{equation}

The EM algorithm alternates between maximising the lower bound with respect to the distribution $q(\mathbf X)$ and the parameter set $\boldsymbol\theta$, respectively, while holding the other fixed. Suppose that the current estimate of the parameter set is $\boldsymbol\theta'$, in the E-step the lower bound is maximised with respect to the distribution $q(\mathbf X)$ while $\boldsymbol\theta'$ is fixed. Since the value of $p(\mathbf Y;\boldsymbol\theta')$ is independent of $q(\mathbf X)$ the maximum of the lower bound occurs when the KL distance disappears or in other words when $q(\mathbf X)$ is exactly the conditional distribution $p(\mathbf X \mid \mathbf Y;\boldsymbol\theta)$. At this point, the lower bound becomes equal to the log-likelihood function. Note in the state-space model the E-step is equivalent to the smoothing problem. The maximum in the M-step is obtained by maximising the lower bound with respect to the parameter set while the distribution $q(\mathbf X)$ is held fixed. Since the distribution $q(\mathbf X)$ is computed using the old parameter estimates and kept fixed during the M-step, it is not equal to $p(\mathbf X \mid \mathbf Y;\boldsymbol\theta)$, resulting in a non-zero KL distance. Therefore, the growth of the log-likelihood function is greater than the growth of the lower bound. Substituting $p(\mathbf X \mid \mathbf Y;\boldsymbol\theta')$ for $q(\mathbf X)$ in \eqref{eq:EM-decomposition1} gives
\begin{align}
 \mathcal{L}(q,\boldsymbol\theta)&=\int_{\mathbf X} p(\mathbf X\mid\mathbf Y;\boldsymbol\theta')\ln p(\mathbf X,\mathbf Y;\boldsymbol\theta)~d\mathbf X-\int_{\mathbf X} p(\mathbf X\mid\mathbf Y;\boldsymbol\theta')\ln p(\mathbf X\mid\mathbf Y;\boldsymbol\theta')~d\mathbf X \nonumber \\
&= \mathbf E_{\boldsymbol\theta'}\left[ \ln p(\mathbf X,\mathbf Y;\boldsymbol\theta)\right] +\mathrm{const.}\nonumber\\
&=\mathcal{Q}(\boldsymbol\theta,\boldsymbol\theta')+\mathrm{const.}~,\label{eq:EM-ComDaLogLike}
\end{align}
where $ \mathbf E_{\boldsymbol \theta'}\left[ \cdot\right] $ denotes expectation taken with respect to the marginal distribution of the states conditioned on the observed field and the current estimate of the parameter set, i.e.  $p(\mathbf X\mid\mathbf Y;\boldsymbol \theta'$). The constant term is the negative entropy of $q(\mathbf X)$ and is independent of $\boldsymbol\theta$.   It becomes clear from \eqref{eq:EM-ComDaLogLike} that, the M-step essentially maximises the expectation of the complete data log-likelihood. 
\section{Calculation of $\boldsymbol\Xi_{0}$ and $\boldsymbol\Xi_{1}$}\label{ap:Xiderivation}
The expressions for $\boldsymbol\Xi_{0}$ and $\boldsymbol\Xi_{1}$ in terms of smoother outputs can be found in \cite{Shumway2000}, here we present the derivation for completeness.  The expression for  $\boldsymbol\Xi_{0}$ as defined in equation~\eqref{eq:Xi0} of the main text is given by
   \begin{equation}\label{eq:appXi0}
	 \boldsymbol\Xi_0=\mathbf E_{\theta'}\left[\sum_{t=0}^{T-1}\mathbf x_t\mathbf x_{t+1}^\top\right].
	\end{equation}                                                                                                
	We start by expanding the equation \eqref{eq:appXi0}, giving
